Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
update missing value labels, special missings, and nabular data

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`. This subtle change is important as it changes the focus from the value of a variable to it's missingness state, making intent clear when performing analysis.

@@

Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing. Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out". The underlying number representation changes from binary to integer. For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example. Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode).

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, and explicitly links missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats that have quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

>>>>

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

### Missing value labels {#missing-labels}

The $R$ matrix is a very common representation of missing values, and can be easily created in the "R" programming language with \code{is.na(data)}. The values 0 and 1 (or equally, FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent - does 0 indicate a the absence of observation, or the presence of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).

@@

Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing. Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out". The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example. Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode). Special missing values do exist in other statistical programming languages, although are limited to 26 or 52 special types of missingness.

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats that have quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
